---
title: "From Cambodia to the Soviet Union: What Factors Influence the Probability of Death for Members of Totalitarian Regimes?"
author: "Katrin Aug, Leonhard Gruber, Samuel Hashem Zehi, Michaela Preclíková, Dominik Ruso"
date: "12/02/2021"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
bibliography: bibs.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\newpage
\section{Introduction}
Dictatorships and totalitarian regimes differ highly with respect to their propagandistic strategies, economic policy, foreign policy, their rhetoric or their treatment of opponents. While some seize power via the corruption of democratic mechanisms like the NSDAP did in 1933, others – like the Military Junta in Argentina or the Libyan dictator Muammar al Gaddafi – do so by using pure military force. Some are even set into power by foreign nations. Once in charge patterns of legitimization need to be constructed. May it be political ideas like communism, religious dogma or a believed racist superiority – totalitarian regimes are manifold.

Despite all differences: What unifies them is the exertion of violence to stay in power, may it be against their own citizens or their neighbors or both. Every authoritarian, dictatorial regime uses collective violence as a tool to exercise control. Hereby dictators throughout history turned out to be very inventive. German concentration camps and Soviet gulags are known worldwide and gained questionable fame. In the last decades newspapers were full of reports about torture chambers in Libya or Syria, China's suppression of their Uyghur minority isn't a secret, too. In Argentina the stories of dissidents being thrown into the sea out of flying helicopters are commonly known. Of course, not every authoritarian regime commits thousandfold mass murder – sometimes locking up journalists and political opponents is enough to remain in power. Nevertheless, very often violence falls back on those who exert it. As long as the existence of totalitarian regimes and dictatorships their members fell victim to violence themselves. An internal fight for power, a lost war or a revolution – there are many situations to think of. Also, there are many circumstances that can be thought of that influence violent death of members of such regimes.
\section{Model Specification}
In this section, we give a short overview on how the model was constructed, as well as brief justifications as to why some variables are included. The variable selection is based on variables of interest, rather than trying to maximize the predictive power of the model. The dummy for the country in which the regime is, or was, located. There might be cultural reasons why countries in Latin America or Asia might differ from countries in Europe. Furthermore, the duration of the regime is included, as it might be the case that regimes that have been in power for longer have an increasing effect on the probability of death for its members, simply due to being longer in power giving more chances to die, either natural or not. Related to this, the demeaned end year of the regime is included, as the possibility of time trends seems plausible. E.g.\ when regimes were overturned 100 years ago, the population might have been less averse to peaceful transitions. An earlier version of this model included the start year, however, the variance inflation factor (VIF) on the coefficient was relatively large, compared to the others, so it was excluded from the model. End-of-regime two-year average GDP growth is included, as a worse economic situation might lead to a more desperate population, which in turn might lead to increased propensity of a more violent approach to overturning the current regime. Lastly, all other variables are included simply out of interest, not due to an assumption on the sign and size of the partial correlations with the outcome variable. 

\section{Results and Interpretation}
As Table 1 shows, quite a lot of coefficients are not statistically significant. Ignoring this for a moment, there is a positive relationship between death during, or at the end of the regime and tenure in regime. The same is observed for regime end: with higher deviation from the mean end year of regime, the log-odds of death increases. The opposite effects are observed for female and GDP growth. Being female decreases the log-odds of death during, or at the end of the regime. Higher GDP growth in the country also decreases the log-odds of death. Even though other variables are insignificant, we can still look at their coefficient signs. Being part of the regime in a European country decreases the log-odds of death. The same applies for economists or individuals who were previously military members. The insignificance may stem from the small sample size. 

Log-odds can show us relationships between the dependent and independent variables, whereas it is preferable to also use odds and marginal effects to understand the coefficients better.
The odds for death are 1.036 times higher when tenure increases by one unit, keeping other variables constant. In other words, the odds for death increase by 3.6\% with one year increase. For a one unit increase in the GDP growth in the last 2 years of the regime we expect to see almost 19% decrease in the odds of dying during the regime, i.e. the chance of not dying is higher for members whose countries experienced a GDP growth. The later the regime ended, compared to the average end year of all included regimes, the higher are the odds of dying during, or at the end of the regime. Even though it is insignificant, being part of a regime in Europe decreases the average probability of death by 10.7 percentage points.

As some examples of what the model would predict, we chose four well known personalities from totalitarian regimes. We start with two members of the NSDAP in Germany. Firstly, Adolf Hitler, who was leading the german regime throughout its entire existence, having a predicted probability to die of 10.44\%. The second member is Ernst Röhm, the co-founder of the \emph{Sturmabteilung}, and early friend of Adolf Hitler. His predicted probability of dying is lower, sitting at only 7.27\%. A fierce rival, at least through some part of the regime, of the German regime was the Soviet Union, and with it its leader Joseph Stalin. Stalin managed to hang on to power for quite a long time, thus having a higher probability of dying. Additionally, the economic growth of the Soviet Union in the years before its collapse was quite grim, also driving up the death probability, all in all, his predicted value is at 32.27\%. However, it should be noted that his death came some time before the fall of the regime, so his influence on the economic situation was quite limited. As a last example, we chose Pol Pot, who governed Cambodia for four years. Him not being from Europe, combined with negative GDP growth leads to a predicted probability of 31.76\%. Noteworthy here is, that Pol Pot did not die in any way related to his regime, but rather died of a heart attack before any punishment for his actions was undertaken. So, while his predicted probability is quite high, at least compared to the generally very low predicted probabilities, the model does not correctly classify him when choosing a cutoff at 50\%. 
\section{Model Diagnostics and Issues}
None of the coefficients exhibit a VIF value that seems to lead to any larger concerns, using a critical value of 10 on the determination for problematic imperfect multicollinearity. Some of the standards appear quite large, this is, at least partly, due to the VIFs, as indicated in Table 2 of Appendix A. As is denoted in Table 1, the McFadden R$^2$ is 0.091, so the model appears to have some explanatory power. To put this in further context, the hit-rate of the model, using a cutoff value of 0.5, is 88.07\%, compared to a hit-rate of 87.66\% when simply guessing the most likely outcome of not dying. One issue with such a model is, per definition, that external validity is not given, as we pre-select for being part of an authoritarian regime. Furthermore, the inclusion of variables specific for the regime, such as the duration, don't allow for out-of-sample validation methods, only allowing for validation and calibration based on random sub-samples of this particular dataset. 
\section{Conclusion}

The model cannot capture a lot of the country-specific and time-specific factors which drive the unobserved dependent variable, at least given the data available. Thus, we are somewhat limited in the conclusions we are able to draw from this model. However, within this model, the first of four main drivers of the variation of the death probability, the economic growth rates immediately before the regime falls, which negatively influences the unobserved outcome variable. Secondly, the year in which the regime fell, which has a positive sign on the coefficient, additionally, the tenure of the individual member in their respective regime, which also has a positive coefficient. Lastly, female members of totalitarian regimes would have a significantly smaller probability of death, compared to their c.p.\ male counterparts.
\newpage
\section*{Appendix A: Tables}
\addcontentsline{toc}{section}{Appendix A: Tables}
\begin{table}[!h] \centering 
  \caption{Logistic Regression Results (Bootstrapped Coefficients and Standard Errors)} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{1}{c}{\textit{Dependent variable:}} &  & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} \cline{4-4} 
\\[-1.8ex] & Death Dummy & & Death Dummy \\ 
\hline \\[-1.8ex] 
 Europe Dummy & $-$1.052 &  Regime Duration & $-$0.001 \\ 
  & (0.626) &   & (0.012) \\ 
  & \\ 
 Female Dummy & $-$15.502 &  Age at Regime Entrance & 0.010 \\ 
  & (0.376) &   & (0.018) \\  
  & \\ 
 Tenure in Regime & 0.036 &  2Y GDP Growth$^1$ & $-$0.208 \\
  & (0.015) &  & (0.061) \\
  & \\ 
 Economist Dummy & $-$2.675 &  Military Dummy & $-$0.186 \\ 
  & (5.511) &   & (0.560) \\ 
  & \\ 
 Regime End & 0.059 &  Constant & $-$2.150 \\
  & (0.024) &   & (1.025) \\ 
  & \\ 
  \hline \\[-1.8ex] 
Observations & 721 & Log Likelihood & $-$245.020 \\ 
McFadden R$^2$ & 0.091 & & \\ 
\hline 
\hline \\[-1.8ex] 
\multicolumn{4}{l}{\textit{Note:} \vtop{\hbox{\strut Bootstrap standard errors of coefficients in round brackets.}}} \\
\multicolumn{4}{l}{\quad\quad\,\,\, \vtop{\hbox{\strut $^1$Two-year GDP growth measured in the last two years of the respective regimes.}}}
\end{tabular} 
\end{table}

\begin{table}[!h] \centering 
  \caption{Variance Inflation Factors} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc lc lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 Europe Dummy & 3.54 &  Regime Duration & 3.97 & Female Dummy & 1.00 \\ 
  & \\ 
 Age at Regime Entrance & 1.22 &  Tenure in Regime & 1.15 & 2Y GDP Growth & 4.19 \\ 
  & \\ 
 Economist Dummy & 1.01 &  Military Dummy & 1.39 & Regime End & 8.01 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

\newpage
\section*{Appendix B: Project Code}
\addcontentsline{toc}{section}{Appendix B: Project Code}
```{r l1, fig.height=5, fig.width=6, message=FALSE, warning=FALSE}
# clear workspace
rm(list = ls())

# load needed libraries
library(readr)
library(pscl)
library(car)
library(pROC)
library(margins)

# read dataset
data <- read_csv("C:/Users/samue/Downloads/Studium/Economics (Master - Vienna)/1. Semester/Microeconometrics/Project/Overview.csv")

# check if import worked
head(data)
head(data$ELITE_NAME)
table(data$ELITE_EXITFATE)
table(data$ELITE_EXITTYPE)


# create dummy for death during regime or at end
dim(data); n <- dim(data)[1]
death1 <- rep(1,n) 

# dummy for first type of death
for(i in 1:n){
  death1[i] <- ifelse(data$ELITE_EXITTYPE[i] != 'Demotion' 
                      && data$ELITE_EXITTYPE[i] != 'Expulsion'
                      && data$ELITE_EXITTYPE[i] != 'Death- natural'
                      && data$ELITE_EXITTYPE[i] != 'Regime Change'
                      && data$ELITE_EXITTYPE[i] != 'Resignation',1,0)}
table(death1)

# dummy for second type of death
death2 <- rep(1,n)
for(i in 1:n){
  death2[i] <- ifelse(data$ELITE_EXITFATE[i] == 'Execution',1,0)
}
table(death2)

# merge dummies
a <- which(death2==1)
b <- which(death1==1)
c <- c(a,b)
death <- rep(0,n)
death[c] <- 1
table(death)

# dummy for country being in europe
EUROPE <- rep(0,n)
for(i in 1:n){
  EUROPE[i] <- ifelse(data$STATE[i] == 'Germany' || data$STATE[i] == 'Poland'
                      || data$STATE[i] == 'East Germany' || data$STATE[i] == 'Hungary'
                      || data$STATE[i] == 'Norway' || data$STATE[i] == 'Romania'
                      || data$STATE[i] == 'Soviet Union',1,0)
}
table(EUROPE)

# dummy for military as occumpation outside of regime
MIL <- rep(0,n)
for(i in 1:n){
  MIL[i] <- ifelse(data$ELITE_OCCUPATION[i] == 'Soldier' 
                   || data$ELITE_OCCUPATION[i] == 'State Security'
                   || data$ELITE_OCCUPATION[i] == 'Army officer'
                   || data$ELITE_OCCUPATION[i] == 'Naval officer'
                   || data$ELITE_OCCUPATION[i] == 'Military Police officer'
                   || data$ELITE_OCCUPATION[i] == 'Police officer'
                   || data$ELITE_OCCUPATION[i] == 'Air Force Officer'
                   || data$ELITE_OCCUPATION[i] == 'Air Force officer',1,0)
}
table(MIL)

# dummy for economists
ECON <- rep(0,n)
for(i in 1:n){
  ECON[i] <- ifelse(data$ELITE_OCCUPATION[i] == 'Economist' 
                    || data$ELITE_OCCUPATION[i] == 'economist',1,0)
}
table(ECON)

# create function to extract date from string
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

# get regime end year
END <- data$REG_END
END <- substrRight(END, 4)
END <- as.numeric(END)
table(END)

# get regime start year
START <- as.numeric(substrRight(data$REG_START, 4))
table(START)

# take care of regimes that have not ended
END <- ifelse(is.na(END),2020,END)
table(END)

# variable for regime duration
DURATION <- END-START
table(DURATION)

# standardize regime start year
mean(START)
START <- START-mean(START)

mean(END)
END <- END - mean(END)

#View(data)



# make data numeric for the model
death <- as.numeric(death)
ELITE_FEMALE <- as.numeric(data$ELITE_FEMALE)
ELITE_ENTERAGE <- as.numeric(data$ELITE_ENTERAGE)
ELITE_RETENURE <- as.numeric(data$ELITE_RETENURE)
EC_GR_2Y <- as.numeric(data$EC_GR_2Y)

# create dataset only based on relevant variables
data0 <- cbind(data$ELITE_NAME,data$STATE,death,EUROPE,START,DURATION,ELITE_FEMALE,
               ELITE_ENTERAGE,ELITE_RETENURE,EC_GR_2Y,ECON,MIL,END)
data <- cbind(death,EUROPE,START,DURATION,ELITE_FEMALE,ELITE_ENTERAGE,
              ELITE_RETENURE,EC_GR_2Y,ECON,MIL,END)





data <- as.data.frame(data)
data0 <- as.data.frame(data0)
#View(data0)

# get vectors for specific people:
hitler <- type.convert(unlist(c(1,data0[21,-c(1:3,5)]),use.names = FALSE),dec='.')
# check if person is correct
data0[21,9] == data[21,7]

# Röhm:
roehm <- type.convert(unlist(c(1,data0[35,-c(1:3,5)]),use.names = FALSE),dec='.')
# Stalin:
stalin <- type.convert(unlist(c(1,data0[799,-c(1:3,5)]),use.names = FALSE),dec='.')
# Pol Pot
pol_pot <- type.convert(unlist(c(1,data0[223,-c(1:3,5)]),use.names = FALSE),dec='.')

# remove NAs
data <- na.exclude(data)
data0 <- na.exclude(data0)


#View(data0)
#View(data)



# logit model
model1 <- glm(death ~ EUROPE+START+DURATION+ELITE_FEMALE+ELITE_ENTERAGE
              +ELITE_RETENURE+EC_GR_2Y+ECON+MIL,
    family = binomial(link = 'logit'), data=data)
summary(model1)

# pseudo R2s
pR2(model1)

# variance inflation factors
vif(model1)

# cor(data) shows correlations across all variables
cor(EUROPE,START)

# model without START
model2 <- glm(death ~ EUROPE+DURATION+ELITE_FEMALE+ELITE_ENTERAGE+ELITE_RETENURE
              +EC_GR_2Y+ECON+MIL+END,
    family = binomial(link = 'logit'), data = data)
summary(model2)
pR2(model2)

# Variance Inflation Factors
vif(model2)

############################################
######### BOOTSTRAP ETIMATION ##############
############################################

set.seed(1234); N <- 500
Coeff_boot <- matrix(rep(0,N*10),ncol = 10, nrow = 500)
for(i in 1:N){
  boot_sample <- sample(1:n,size = n,replace = TRUE)
  mod_boot <- glm(death ~ EUROPE+DURATION+ELITE_FEMALE+ELITE_ENTERAGE+ELITE_RETENURE
              +EC_GR_2Y+ECON+MIL+END,
    family = binomial(link = 'logit'), data = data[boot_sample,])
  Coeff_boot[i,] <- mod_boot$coefficients
}
coeff_boot <- apply(Coeff_boot,2,mean)
se_boot <- rep(0,10)
for(i in 1:10){
  se_boot[i] <- sqrt(var(Coeff_boot[,i]))
}
txt <- c('Intercept','EUROPE','DURATION','FEMALE','AGE','TENURE',
         '2YGR','ECON','MIL','END')
rbind(txt,round(coeff_boot,3),round(se_boot,3))


# get hit rate of the model with a 0.5 cutoff (based on original model)
cutoff <- 0.5
# get predicted probabilities
latent_pred <- predict.glm(model2,type = 'response')
# get binary result from the cutoff
latent_bin <- ifelse(latent_pred >= cutoff,1,0)
# hitrate 
mean(latent_bin==data$death)
# percentage by just guessing more likely outcome
1-mean(data$death)

# maximal predicted probability
max(latent_pred)
densityPlot(latent_pred, ylab = 'Density', 
            xlab = 'Predicted Probability',
            main = 'Density Plot of Latent Probability of Death')


# check coefficient vector lengths
length(model2$coefficients);length(hitler)
# check vectors for matching variables
hitler
model2$coefficients
# get probabilities
coeff_boot <- type.convert(coeff_boot,dec = '.')
Pr_hitler <- 1/(1+exp(-t(coeff_boot)%*%(hitler)))
Pr_hitler
Pr_pol_pot <- 1/(1+exp(-t(coeff_boot)%*%(pol_pot)))
Pr_pol_pot
Pr_roehm <- 1/(1+exp(-t(coeff_boot)%*%(roehm)))
Pr_roehm
Pr_stalin <- 1/(1+exp(-t(coeff_boot)%*%(stalin)))
Pr_stalin




# check for percentage of economists and military
mean(na.omit(data$ECON)); mean(na.omit(data$MIL))

mean(na.omit(data$death))

# average marginal effects
mar <- margins(model2)
summary(mar)
mean(na.omit(data$ELITE_FEMALE))
mean(na.omit(data$death[na.omit(data$ELITE_FEMALE) == 1]))
# turns out no women died
```

\newpage
\section*{References}
\addcontentsline{toc}{section}{References}